# Function Exercises - Answers

---

## Preparation (Test Data)

We will use the following DNA sequences as example data.

```{r}
dna1 <- c("A","G","C","T","G")
dna2 <- c("A","T","A","T","A","T")
dna3 <- c("G","G","C","C","G","G")
dna4 <- c("N","N","A","T","G")
dna5 <- c("A","G","X","T")   # unknown character test

dna_list <- list(dna1, dna2, dna3, dna4)
```

## Level 1 — Easy

### Exercise 1: `square_value(x)`

**Expected output:** `square_value(4) = 16`

```{r}
square_value <- function(x) {
  # x: number given to the function

  result <- x * x
  # result: the square of x (x multiplied by x)

  return(result)
  # return(): specifies the value that the function sends back
}

square_value(4)
```

**What did we learn?**

- A function receives input (`x`)
- It performs a calculation
- It returns the result using `return()`

---

### Exercise 2: `dna_length(dna)`

**Expected output:** `dna_length(dna1) = 5`

```{r}
dna_length <- function(dna) {
  # dna: a vector containing bases like A, T, G, C

  length_value <- length(dna)
  # length(): returns how many elements are in the vector

  return(length_value)
}

dna_length(dna1)
```

**What did we learn?**

- `length()` returns the size (number of elements) of a vector.

---

## Level 2 — Medium (for + if + counter)

### Exercise 3: `gc_count(dna)`

**Expected output:** `gc_count(dna1) = 3`

```{r}
gc_count <- function(dna) {

  GC_total <- 0
  # counter: starts at 0 and increases when we find G or C

  for (n in dna) {
    # for loop: goes through each base in the DNA sequence
    # n: the current base (for example "A" or "G")

    if (n == "G" || n == "C") {
      # if condition: checks whether the base is G or C
      # || means "OR"

      GC_total <- GC_total + 1
      # if G or C is found, increase the counter by 1
    }
  }

  return(GC_total)
  # return the total number of G and C bases
}

gc_count(dna1)
```
**What did we learn?**

- Counter logic: start with `0`, increase with `+1` when the condition is true  
- Using a `for` loop to iterate over elements  
- Using `if` to check conditions  

---

### Exercise 4: `gc_ratio(dna)`

**Expected output:** `gc_ratio(c("G","C","A","T")) = 0.5`

```{r}
gc_ratio <- function(dna) {

  GC_total <- 0

  for (n in dna) {
    if (n == "G" || n == "C") {
      GC_total <- GC_total + 1
    }
  }

  ratio <- GC_total / length(dna)
  # GC ratio = (number of G + C) / (total number of bases)

  return(ratio)
}

gc_ratio(c("G","C","A","T"))
```

**What did we learn?**

- To calculate a ratio we use: **part / total**

---

## Level 3 — Medium / Advanced (Classification)

### Exercise 5: `gc_status(dna)`

**Expected output:** `gc_status(dna3) = "GC-rich"`

```{r}
gc_status <- function(dna) {

  ratio <- gc_ratio(dna)
  # instead of writing the loop again we call the gc_ratio() function

  if (ratio > 0.5) {
    # if ratio is greater than 0.5, GC content is high
    return("GC-rich")
  } else {
    return("GC-poor")
  }
}

gc_status(dna3)
```
**What did we learn?**

- Calling another function inside a function keeps the code shorter and cleaner.

---

## Level 4 — Hard (Input checking + QC)

### Exercise 6: `dna_check(dna)`

**Expected output**

`dna_check(dna1) = "OK"`

`dna_check(dna5) = "ERROR: Unknown base"`

```{r}
dna_check <- function(dna) {

  error_exists <- FALSE
  # logical variable (boolean)
  # FALSE = no error, TRUE = error exists

  for (n in dna) {

    if (n != "A" && n != "T" && n != "G" && n != "C" && n != "N") {
      # if n is not A AND not T AND not G AND not C AND not N
      # then it means there is an invalid character
      # && means "AND"

      error_exists <- TRUE
    }
  }

  if (error_exists == TRUE) {
    return("ERROR: Unknown base")
  } else {
    return("OK")
  }
}

dna_check(dna5)
dna_check(dna1)
```
**What did we learn?**

- Data validation: checking if the input is correct  
- Using `TRUE` / `FALSE` for logical checks  

---

### Exercise 7: `dna_quality(dna)`

**Expected output:** `dna_quality(dna4) = "Uncertain DNA"`

```{r}
dna_quality <- function(dna) {

  N_exists <- FALSE
  # if N appears we consider the DNA uncertain

  GC_total <- 0
  # GC counter

  for (n in dna) {

    if (n == "N") {
      N_exists <- TRUE
      # if we see N anywhere -> TRUE
    }

    if (n == "G" || n == "C") {
      GC_total <- GC_total + 1
    }
  }

  if (N_exists == TRUE) {
    # first check N
    return("Uncertain DNA")
  } else {

    ratio <- GC_total / length(dna)

    if (ratio >= 0.6) {
      return("High GC")
    } else {
      return("Normal DNA")
    }
  }
}

dna_quality(dna4)
```
**What did we learn?**

- First check **critical conditions** (is there an `N`?)
- Then check **other criteria** (GC ratio)

---

### Exercise 8: `list_report(dna_list)`

This function prints a report for each DNA sequence in the list.

**Example output**

1. DNA -> Normal DNA | GC: 0.6
2. DNA -> Normal DNA | GC: 0
3. DNA -> High GC    | GC: 1
4. DNA -> Uncertain DNA

```{r}
list_report <- function(dna_list) {

  idx <- 1
  # idx: counter to show which DNA sequence we are processing

  for (dna in dna_list) {

    check <- dna_check(dna)
    # first check if the DNA has invalid characters

    if (check != "OK") {

      print(paste(idx, ". DNA ->", check))

    } else {

      status <- dna_quality(dna)

      if (status == "Uncertain DNA") {

        print(paste(idx, ". DNA ->", status))

      } else {

        ratio <- gc_ratio(dna)

        print(paste(idx, ". DNA ->", status, "| GC:", ratio))
      }
    }

    idx <- idx + 1
  }
}

list_report(dna_list)
```


**What did we learn?**

- Iterating through a list with `for`
- Calling functions for each element
- Combining text using `paste()`

---

### Exercise 9: `dna_summary(dna)`

Returns a short summary text for one DNA sequence.

**Expected output**

`dna_summary(dna1) = "OK | Normal DNA | GC=0.6"`

```{r}
dna_summary <- function(dna) {

  check <- dna_check(dna)

  if (check != "OK") {

    return(check)

  } else {

    status <- dna_quality(dna)

    if (status == "Uncertain DNA") {

      return(paste("OK |", status))

    } else {

      ratio <- gc_ratio(dna)

      return(paste("OK |", status, "| GC=", ratio))
    }
  }
}

dna_summary(dna1)
dna_summary(dna2)
dna_summary(dna3)
```

**What did we learn?**

- Pipeline logic: **check → quality → (if needed) GC ratio**
- Creating a single summary output string